---
title: "Data visualization"
author: "Michelle Lam"
date: "2022-08-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(janitor)

library(patchwork)
library(ggrepel)
library(gghighlight)
library(paletteer)
library(ggExtra)
library(ggbeeswarm)

library(gapminder)

library(sf)
```

### Read in lizard data

```{r}
lizards <- read_csv(here("data_tidy", "lizards.csv"))
```

```{r}
ggplot(data = lizards) +
  geom_point(aes(x = total_length, y = weight))
```

Create a histogram of total lengths for all lizards in the dataset.
```{r}
plot_1 <- ggplot(data = lizards, aes(x = total_length)) +
  geom_histogram(color = "orange",
                 fill = "purple",
                 size = 0.3, 
                 linetype = "dotted")

plot_1
# use aes when you want to map an aesthetic onto a variable
# if you refer to a constant within aes it will show up in the legend. constants should not be within aes
```

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(
    shape = 22,
    color = "cyan4", 
    fill = "yellow",
    size = 4,
    alpha = 0.4
  )
```

```{r}
ggplot(data = lizards, 
       aes(x = total_length, y = weight)) +
  geom_point(
    aes(color = common_name, 
        size = total_length),
    alpha = 0.5) +
  theme_minimal()
```

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  facet_wrap(~common_name, scales = "free")

# R automtaically scales so they all have the same x and y axis differences, to change the scale use scales = free
```

```{r}
lizard_counts <- lizards |> 
  group_by(common_name) |> 
  summarize(
    counts = n()
  )
# n counts number of rows by group, if you want to add multiple columns in grouped variable use group_by and summarize

lizard_counts <- lizards |> 
  count(common_name)
```

```{r}
ggplot(data = lizard_counts, 
       aes(y = fct_reorder(common_name, n), 
           x = n)) +
  geom_col()
# default for character variable is to plot in alphabetical order. can't specify order of a character, can specify order of a factor
# fct_reorder allows you to reorder the common_name by the n variable
```

Create scatterplots, faceted by species, for total length versus weight. The color and shape of points should differ based on the species (common name). Change the theme. Make each facted panel have flexible axis limits (scales should be free).
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name,
                 shape = common_name),
             alpha = 0.6, size = 2) +
  theme_minimal() +
  facet_wrap(~common_name, scales = "free") +
  labs(x = "Total length (mm)", 
       y = "Weight (grams)",
       title = "Jornada Basin lizard sizes",
       subtitle = "All lizards follow the standard length-weight relationship, with Western Whiptails being largest",
       caption = "Data: Jornada Basin LTER, CCBY")
```

### Fun with scales

```{r}
# scale_color_gradient specifying low value and high value
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_gradient(low = "purple", high = "orange")

# scale_color_gradientn specify array of colors
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_gradientn(colors = c("magenta", "cyan4", "darkorchid3", "dodgerblue", "green", "yellow"))

# scale_color_steps: set low value and high value and it will bin the values in the gradient
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_steps(low = "red", high = "black")

# scale_color_steps2: give the low, midpoint, and high values and set colors on the gradient
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = total_length)) +
  scale_color_steps2(low = "purple",
                     mid = "white",
                     high = "orange",
                     midpoint = 150,
                     breaks = c(50, 75, 150, 180, 220, 280))
# scale_color_stepsn for greater than three steps
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = total_length)) +
  scale_color_stepsn(colors = c("orange", "red", "purple"),
                     breaks = seq(from = 0, 
                                  to = 300, 
                                  by = 60))
```

Make a horizontal boxplot with common name on the y-axis, total length on the x-axis, with boxplot fill chaning based on the common name
```{r}
# create new data set, recast common name from character to factor, fct_reorder put them in order based on median of total_length based of each common_species group
# character vectors can't assign an order, but factors you can
# to reverse factor order use rev() before fct_reorder
lizards_median <- lizards |> 
  mutate(common_name = fct_reorder(common_name, total_length, .fun = median))

ggplot(data = lizards, aes(x = total_length, y = common_name)) +
  geom_boxplot(aes(fill = common_name))

ggplot(data = lizards_median, aes(y = common_name, x = total_length)) +
  geom_boxplot(aes(fill = common_name), show.legend = FALSE) +
  scale_fill_paletteer_d(palette = "palettetown::squirtle") +
  theme_minimal()

# to view palettes type View(palettes_d_names) in console
```

```{r}
# highest resolution that you specify will be applied to the graph
# within theme change gridlines, font, axis text
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  theme(panel.grid.major.x = element_line(color = "red"),
        panel.grid.minor.y = element_blank(),
        axis.title = element_text(color = "purple", size = 10),
        axis.text.y = element_text(color = "green"),
        text = element_text(size = 18),
        panel.background = element_rect(color = "purple", fill = "yellow")) +
  annotate("text", x = 300, y = 50, label = "Wooooo\ntest", color = "blue") +
  geom_vline(xintercept = 250, linetype = "dashed", color = "orange", size = 4)
```

Repulsive labels with `ggrepel`

```{r}
wws_lizards <- lizards |> 
  filter(common_name == "western whiptail",
         site == "sand")

# repel moves labels so they don't overlap the points and creates lines from labels to data points 
ggplot(data = wws_lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  geom_text_repel(aes(label = toe_num))

gapminder |> 
  filter(year == 2002, 
         continent == "Europe") |> 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_text_repel(aes(label = country), size = 3)
```

gghighlight: applies highlight to points based on condition set
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  gghighlight(toe_num == 250, label_key = toe_num) +
  gghighlight(weight >30, label_key = toe_num)
```

### Make a map!
```{r}
# read in the spatial data file
jornada_veg <- read_sf(here("data_raw", "spatial_vegetation", "doc.kml"))

ggplot() +
  geom_sf(data = jornada_veg, 
          aes(fill = Name),
          color = NA) +
  theme_minimal() +
  scale_fill_paletteer_d(palette = "ggthemes::manyeys")
```

